Asymmetric relationships and their effects on coexistence

### 2023-07-13 paper/RefereeResponses1.tex

\noindent ***\emph{We address your questions and comments below.} \\

\noindent C4) My biggest question is related to the removal of ATAs from a given true bivariate distribution.  This is described in SI Section S2 and shown visually for an example in SI Figure S1.

My first comment related to this is that some form of this text (whcih is relatively short---around half a page in the SI) and, importantly, the Figure should be in the main text.  I think this is such a key part of the definitions that are being used for the rest of the paper that readers will want to get an intuition for what the "partial sharp" distributions look like and how they are arrived at. \\

\noindent ***\emph{We have done what the referee requests, here, and we agree with this request. Thanks for
making the request. This request is somewhat in tension with other requests, from other referees, 
to minimize the quantitative 
content of the main text, and to keep the overall length under the journal limit while adding a bunch of
verbal explanations requested by the other referees. We point this tension out here, and refer back to it in
other responses, below. We also point out that we have not only moved the figure back to the main text,
we have also added a panel to the figure in response to the related comment CX, below.} \\
%DAN: Waiting on Jasmin to add the new panel. In other replies, you can refer to this reply to justify when you
%decide to add explanations to the SI instead of to the main text, and when you have to respond to requests
%to minimize the quant. Also fill in the correct value of CX above.

@@

\noindent C5) Perhaps equally important to clarifying this process for the readership is whether this process is unique. I could not quite get a sense of whether there could be alternative versions of this removal of ATA from a bivariate distribution, that still satisfy the authors' constraints on marginal distributions and overall correlation between the two variables. Are there alternatives, or is the authors' procedure unique in achieving the desired outcome?  If the former, then I would want to understand if the outcomes downstream tend to be robust to the specific method used. If the latter, it would be helpful to have an explanation as to why this is a unique decomposition. \\

\noindent ***\emph{This is a good question, thanks for asking it. The extent to which the process is 
unique is now covered in a new SI section, refered to from the main text at the end of section 2.3, 
right at the end of the text the referee asked us to move to the main text. There are some choices involved,
but they are pretty cannonical. See the new SI section for explanation. } \\
%Dan to do
%<Address the issue of whether and to what extent the process of removing ATAs is unique – probably just with a one-sentence callout in the main text, and some material in the SM (possibly a new section)>

### 2023-07-13 paper/Paper.Rnw
working on replies - moved fig S1 back to main text

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of 
tail association are removed -- see SI section \ref{SI-sect:remove_asymmetries} for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$. We will refer to this term as 
the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
Note that brackets, $[\cdot]$, and parentheses, $(\cdot)$, are used distinctly in our notation,
throughout the paper - they are not interchangeable.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

@@

>>>>

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of 
tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$. We will refer to this term as 
the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
Note that brackets, $[\cdot]$, and parentheses, $(\cdot)$, are used distinctly in our notation,
throughout the paper - they are not interchangeable.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? Fig. \ref{fig:pedagog2} provides an illustration of the steps we describe here. 
We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal (Fig. \ref{fig:pedagog2}a,b).  
Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$ (Fig. \ref{fig:pedagog2}c). We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$ (Fig. \ref{fig:pedagog2}d). 
The value of $\ddot{\rho}_i$ is then determined in one of 
four alternative ways: either by setting $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created similarly.  
For comparison, Fig. \ref{fig:pedagog2} also shows, for the example explored by that figure, 
the distribution $(E_i^\shp,C_{i \bs i}^\shp)$
defined in the previous section (following \cite{Ellner_2016}).
%DAN: Jasmin will be adding that panel.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/theory_fig_col.pdf}
\caption[Visual guide to removing asymmetries of association]{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{\nat}$ and  $C_{i \bs i}^{\nat}$,
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
Panel (e) show the distribution $(E_i^\shp,C_{i \bs i}^\shp)$, defined in Theory
following \citep{Ellner_2016}, for comparison.
%DAN: jasmin is adding this new panel
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}
